Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies

### 2019-02-06 man/ms.Rmd
update

Drought is a natural phenomenon occurring when water availability is significantly below-normal levels [@Dai2011]. It is one of the most frequent and widespread climatic extremes affecting ecological systems across every climatic zone worldwide [@Dai2011; @Ummenhofer2017]. Despite drought is considered a feature of Mediterranean climate [@Lionello2012], an increase in the frequency and severity of drought events has been recorded in the last decades [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported [@Spinoni2017b]. This is especially important, given the climate change projections for the Mediterranean area that forecast that climate extreme events, like droughts, will become more frequent and severe [@IPCC2013; @Trenberth2014; @Spinoni2017a].

@@

Ecosystem responses to drought are influenced by other drivers of environmental changes [@Fischer2006; @Oliver2014; @Penuelas2017]. In a global-change context, the impacts of drought on vegetation need to be evaluate considering the interactions with other drivers of change [@Clavero2011; @DoblasMiranda2017]. This is especially relevant for areas with a long history of landscape modification, such us Mediterranean region, where land-use change is a key driver [@Navarro2013; @Ameztegui2016], considering also that the interactions of these two factors, namely drought and land-use change, are crucial for Mediterranean forests [@DoblasMiranda2017].

@@

Several studies analyzed the drought effects on Mediterranean tree species representing the southernmost limit of their distribution
[*e.g.*, @SanchezSalguero2012; @Linares2014; @Dorado2017b; @Caminero2018]. Some of these studies focused on the resilience to drought of rear-edge populations [*e.g.*, @Herrero2014]. The assessment of resilience --*i.e.* the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance [@Holling1973; @Hodgson2015]-- to climate disturbances provides critical information about the capacity of the forests to maintain their structure and to continue providing valuable ecosystem services. This is particularly relevant for populations located in the rear-edge of their current distribution, especially when they are located in Mediterranean mountains.

Here we assessed the resilience to drought of rear-edge populations of a tree Mediterranean species combining several approaches: remote sensing and dendroecologial estimates of growth. Remote sensing can be used for studying droughts impacts on ecosystems [*e.g.*, @Deshayes2006; @Zhang2013; @AghaKouchak2015; @McDowell2015; @Norman2016]. Vegetation indices obtained from satellite, such us EVI (Enhanced Vegetation Index) are used as a proxy for assessing vegetation functioning and aboveground net primary production [@Huete2002], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013; @Vicca2016]. Drought impacts on vegetation using remote sensing have been analyzed at different scales. For instance, a global-scale study showed a stronger influence of the drought for drier biomes [@VicenteSerrano2013]. For Mediterranean, studies at regional and local scales found a large spatial and seasonal variability of the effects of drought on vegetation [@VicenteSerrano2007; @Gouveia2017].

Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality [@Fritts1976; @Dobbertin2005; @Bhuyan2017] and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976].

The combination of both approaches, namely remote sensing and dendroecology, has been used to assess the effects of droughts on vegetation along ecological or elevational gradients [*e.g.*, @Dorman2013; @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018]

Several studies combined the use of satellite information with dendroecological data to assess the effect of droughts on vegetation [@Dorman2013; @VicenteSerrano2013; @Sanguesa2014; @Coulthard2017]; and recently, @Gazol2018 evaluated the resilience of several forests in Spain combining information derived from remote sensing and tree-ring data.

@@

The forests of this species reaches their `REVIEW` southernmost European limit in Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations between 860 m and 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August). There are eight Pyrenean oak patches (2400 Has) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys. Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010] and these populations are considered as a rear edge of the habitat distribution, which is important in determining habitat responses to expected climate change [@Hampe2005]. For this reason, the populations of Pyrenean oak forests at Sierra Nevada are considered relict forests [@Melendo2000; @Vivero2000], having these forests undergone intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2010; @Valbuena2017]. Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" [@Vivero2000] and they are expected to suffer the impact of climate change, reducting its suitable habitats in the rear-edge due to their greater hydric requirements compared to other more drought-tolerant Mediterranean evergreen oak and pine species [@GeaIzquierdo2013].`REVIEW CITA`

>>>>

Drought is a natural phenomenon occurring when water availability is significantly below-normal levels [@Dai2011]. It is one of the most frequent and widespread climatic extremes affecting ecological systems across every climatic zone worldwide [@Dai2011; @Ummenhofer2017]. Despite drought is considered a feature of Mediterranean climate [@Lionello2012], an increase in the frequency and severity of drought events has been recorded in the last decades [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported [@Spinoni2017b]. This is especially important, since the climate change projections for the Mediterranean area forecast that climate extreme events, like droughts, will become more frequent and severe [@IPCC2013; @Trenberth2014; @Spinoni2017a].

@@

Ecosystem responses to drought are influenced by other drivers of environmental changes [@Fischer2006; @Oliver2014; @Penuelas2017]. In a global-change context, the impacts of drought on vegetation need to be evaluate considering the interactions with other drivers of change [@Clavero2011; @DoblasMiranda2017]. This is especially relevant for areas with a long history of landscape modification, such us Mediterranean region, where land-use change plays a significant role [@Navarro2013; @Ameztegui2016], considering also that the interactions of these two factors, namely drought and land-use change, are crucial for Mediterranean forests [@DoblasMiranda2017].

@@

Several studies analyzed the drought effects on Mediterranean tree species representing the southernmost limit of their distribution
[*e.g.*, @SanchezSalguero2012; @Linares2014; @Dorado2017b; @Caminero2018]. Some of these studies focused on the resilience to drought of rear-edge populations [*e.g.*, @Herrero2014]. Resilience is the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance [@Holling1973; @Hodgson2015]. The assessment of resilience to climate disturbances provides critical information about the capacity of the forests to maintain their structure and to continue providing valuable ecosystem services. This is particularly relevant for populations located in the rear-edge of their current distribution, especially when they are located in Mediterranean mountains.

Here we assessed the resilience to drought of rear-edge populations of a tree Mediterranean species combining several approaches: remote sensing and dendroecologial estimates of growth. Remote sensing can be used for studying droughts impacts on ecosystems [*e.g.*, @Deshayes2006; @Zhang2013; @AghaKouchak2015; @McDowell2015; @Norman2016]. Vegetation indices obtained from satellite, such us EVI (Enhanced Vegetation Index) are used as a proxy for assessing vegetation functioning and aboveground net primary production [@Huete2002], and changes in the patterns of these indices are considered as an indicator of the forest response to external environmental factors, such us droughts [*e.g.* @VicenteSerrano2013; @Vicca2016]. Drought impacts on vegetation using remote sensing have been analyzed at different scales. For instance, a global-scale study showed a stronger influence of the drought for drier biomes [@VicenteSerrano2013]. For Mediterranean area, studies at regional and local scales found a large spatial and seasonal variability of the effects of drought on vegetation [@VicenteSerrano2007; @Gouveia2017].

Dendroecological estimates of growth (*i.e.* tree-ring width) are a widely used proxy to study tree vitality [@Fritts1976; @Dobbertin2005; @Bhuyan2017] and the analysis of annual-tree ring widths can be used to study changes in growth as response to drought at the individiual tree-scale [@Fritts1976]. The combination of both approaches (remote sensing and dendroecology), has been used to assess the effects of droughts on vegetation along ecological or elevational gradients [*e.g.*, @Dorman2013; @VicenteSerrano2013; @Coulthard2017], and also for the analysis of growth resilience to drought on several tree species [@Gazol2018]

@@

The forests of this species reaches their southernmost European limit in Sierra Nevada, a high-mountain range located in southern Spain (37°N, 3°W) with elevations between 860 m and 3482 m *a.s.l.*. The climate is Mediterranean, characterized by cold winters and hot summers, with pronounced summer drought (July-August). There are eight Pyrenean oak patches (2400 Has) identified in this mountain range (Figure 1), ranging between 1100 and 2000 m *a.s.l.* and often associated to major river valleys. Sierra Nevada is considered a glacial refugia for deciduous *Quercus* species [@Brewer2002; @Olalde2002; @RodriguezSanchez2010] and these populations are considered as a rear edge of the habitat distribution, which is important in determining habitat responses to expected climate change [@Hampe2005]. For this reason, the populations of Pyrenean oak forests at Sierra Nevada are considered relict forests [@Melendo2000; @Vivero2000], having these forests undergone intensive anthropic use in history [@CamachoOlmedo2002; @Valbuena2010; @Valbuena2017]. Furthermore, conservation status of this species for southern Spain is considered "Vulnerable" [@Vivero2000] and they are expected to suffer the impact of climate change, reducting its suitable habitats in the rear-edge due to their greater hydric requirements compared to other more drought-tolerant Mediterranean evergreen oak and pine species [@GeaIzquierdo2013].
